High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-04-05 PNAS_manuscript/PNAS_manuscript.tex

\significancestatement{Acquiring high-fidelity ancient magnetic field intensity records from rocks is crucial for constraining the long-term evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alteration or non-ideal behavior. We use anorthosites that formed in the deep crust and were brought to the near surface where they acquired thermal remanent magnetizations. These anorthosites have experienced minimal post-formation alteration and yield high-quality paleointensity estimates. In contrast to scenarios of a progressively decaying field leading up to late nucleation of Earth's inner core, these data record a strong field 1.1 Ga. A strong field that persisted over a 14-million-year interval indicates the existence of appreciable power sources for Earth's dynamo during the late Mesoproterozoic.}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy due to the exclusion of light elements \cite{Buffett2000a}. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago \cite{Selkin2007a, Biggin2011a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimated thermal conductivity values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, others have implied higher thermal conductivity values and an age for the inner core that is less than 1.5 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b, Frost2022a}, with some suggesting even younger ages ($<$700 Ma; \citealp{Labrosse2015a, Ohta2016a, Pozzo2022a}). The possibility of late inner core nucleation has motivated proposals of novel power sources to sustain the geomagnetic field through early Earth history including precipitation of light materials such as MgO \cite{Badro2016a, ORourke2016a, ORourke2016b} and SiO$_2$ \cite{Mittal2020a} at the core-mantle boundary. Given that estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of Earth's core and mantle.

@@

Determinations of the absolute value of ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain the record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluid flow. However, intraplate magmatism associated with large igneous provinces is typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift (Fig. \ref{fig:Geologic_map}) is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma during which there were pulsed intervals of more rapid activity \cite{Swanson-Hysell2021a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior far from the continental margin and subsequent orogenesis. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and paleomagnetic data from rift rocks form a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it provides an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have thwarted the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural remanent magnetization (NRM) is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior due to the presence of nonuniformly magnetized grains (with either multidomain \cite{Dunlop2001a} or vortex states \cite{Tauxe2020a}) is sagging or double-slopes as visualized in Arai plots that show thermal remanent magnetization (TRM) acquired versus NRM lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion of the experiment and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Such non-ideal results were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift. While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions \cite{Swanson-Hysell2021a}. The ca. 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7$\pm$0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can protect magnetic inclusions from alteration. In addition, the weathering of the plagioclase crystals do not resulting in the formation of secondary iron oxides, whereas other minerals such as olivine and pyroxene are more likely to alter and influence the magnetic mineralogy. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They were likely heated to tholeiitic magma temperatures and partially recrystallized \cite{Zhang2021b}. Fe-Ti oxides within the plagioclase exsolved above magnetite Curie temperature ($\sim$580\textdegree C; \citealp{Bian2021a}) and subsequently cooled and acquired thermal remanent magnetizations in conjunction with the host diabase at a paleolatitude of 22 $\pm$ 2\textdegree (calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a}). The details regarding the thermal history of these anorthosite xenoliths are discussed in \cite{Zhang2021b}. Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. The anorthosite xenoliths show low anisotropy of thermal remanent magnetization (TRM) acquisition and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows that the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals without strong preferred orientations. Thermal modeling results and paleomagnetic directional data show that the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths to have dominantly single-component magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with remanence being held by low-titanium titanomagnetite \cite{Zhang2021b}.

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). Plagioclase crystals within coarse-grained intrusions often contain abundant elongate Fe-Ti oxide inclusions that are visible with optical microscopy \cite{Feinberg2006a, Wenk2011a, Ageeva2016a}. The Beaver River anorthosite xenoliths lack such large (10s of $\mu$m) oxide inclusions such that they are not visible within the plagioclase crystals using optical microscopy (Fig. \ref{fig:Petro_QDM}C). Magnetic imaging using a quantum diamond microscope (QDM), however, shows that there are magnetic remanence carriers within the plagioclase crystals (Fig. \ref{fig:Petro_QDM}E, G). The sources of these magnetic remanence carriers are likely to be Fe-Ti oxides that exsolved within plagioclase crystals above the Curie temperature of magnetite \cite{Zhang2021b, Bian2021a}. To highlight these distinctive aspects of the Beaver River anorthosite xenoliths, we present petrographic and magnetic imaging data from the Beaver River anorthosite as well as an anorthosite sample from the Duluth Complex Anorthositic Series---an older intrusive complex within the Midcontinent Rift that was not targeted for paleointensity experiments in this study (Fig. \ref{fig:Petro_QDM}). In contrast to the Beaver River anorthosite, the plagioclase crystals of the Duluth Complex anorthosites typically develop interlocking textures that display pronounced igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles within the Duluth Complex plagioclase grains that are typically aligned with the [001] axes of the crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these magnetite needles have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}F, H). As a result of this shape anisotropy, remanent magnetizations are acquired at angles highly oblique to applied fields (Figure \ref{fig:Petro_QDM}F). In an experiment where a second orthogonal field was applied, the first isothermal remanent magnetizations (IRM) of a set of magnetite needles within a plagioclase grain either fail to rotate or flip by 180\textdegree. When they flip, they remain in a direction that is oblique to the applied field direction dictated by shape anisotropy (Fig. \ref{fig:Petro_QDM}F, H). These experiments enable novel visualization of the grain-scale magnetic anisotropy of elongated exsolved (titano)magnetite inclusions within plagioclase grains that leads to magnetic anisotropy observed at the bulk sample scale \cite{Selkin2000a, Feinberg2006a}. In contrast, the remanent magnetizations of the ferromagnetic grains imaged in the Beaver River anorthosite xenoliths, targeted for paleointensity in this study, align with the applied field directions indicating minimal remanence anisotropy  (Fig. \ref{fig:Petro_QDM}E, G). The relative lack of petrofabrics and minimal grain-scale magnetic anisotropy make the Beaver River anorthosite xenoliths a particularly compelling target for paleointensity experiments.

@@

Following IZZI-style paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). Seven anorthosite sites and no diabase sites have specimen results that pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1) where each site represents an individual anorthosite xenolith. The paleointensity quality index (Q$_{PI}$; \citealp{Biggin2014a}) for the anorthosite xenolith sites are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from the anorthosite xenoliths have a mean of 38.28$\pm$11.92 $\mu$T. The site-mean virtual dipole moment is $\sim$82 ZAm$^2$ ($10^{21} Am^2$) at ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthref.org/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we applied an independent statistical method from ref. \citealp{Cych2021a} to all experimental data regardless of their NRM/TRM plot statistics to make a bias-corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on the assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indicate that the anorthosite xenoliths record a high geomagnetic field at ca. 1092 Ma.

\subsection*{Rock magnetism}
\subsubsection*{Coercivity}
Additional rock magnetic data indicate that those anorthosite specimens which pass the paleointensity selection criteria contain dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection along with all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show that both the diabase and anorthosite contain (titano)magnetite as evidenced by the presence of the Verwey transition (Fig. S2;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti contents \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are trending toward lower temperatures (Fig. S2), indicating that their magnetic grains either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}. Another difference is that samples which pass paleointensity selection have distinctly higher average median destructive field values than other anorthosite and diabase specimens (Fig. \ref{fig:coercivity}). Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites having successful paleointensity results can contain magnetic grain populations with peak coercivities ranging from $\sim$44-144 mT with a median of 53 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase specimens tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that magnetic grain populations with more multidomain-like behavior are responsible for the non-ideal paleointensity behaviors during experiments of such specimens \cite{Xu2004a}.

@@

\subsection*{Considering secular variation and cooling rate}
To best characterize the geomagnetic axial dipole field intensity during a certain time period a paleointensity dataset should cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged. Thermal modeling results from ref. \citealp{Zhang2021b} indicate that the Beaver River anorthosite xenoliths were heated to tholeiitic magma temperature ($\sim$1100\textdegree C)---lower than the melting temperature of plagioclase of An$_70$ \cite{Morrison1983a, Doyle2016a} and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a thousand years, partially averaging secular variation within single sites. Another consequence of the anorthosite xenoliths having slowly cooled in the interior of thick diabase intrusions is the bias of paleointensity estimates toward higher values. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. To fully account for the range of various lower temperature bounds used for optimize the paleointensity estimate and the effect of lowering unblocking temperature due to slow cooling, we estimate the anorthosite cooled over a time period on the order of $\sim$10 kyr (from $\sim$580\textdegree C to as low as $\sim$400\textdegree C) based the thermal history model of ref. \citealp{Zhang2021b} and cooling rate correction model of \cite{Pullaiah1975a}. This cooling time corresponds to a cooling rate of $\sim2.5\times10^{-10}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Pullaiah1975a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. This factor may result in an over-correction, as some multidomain grains and pseudo-single domain grains that do not tend to severely bias paleointensity estimates toward higher \cite{Biggin2013a}, may exist in the anorthosite xenoliths as is evidenced through the slightly zigzagging Arai plots (Figure \ref{fig:IZZI_examples}). The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

>>>>

\significancestatement{Acquiring high-fidelity ancient magnetic field intensity records from rocks is crucial for constraining the long-term evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alteration or non-ideal behavior. We use anorthosites that formed in the deep crust and were brought to the near surface where they acquired thermal remanent magnetizations. These anorthosites have experienced minimal post-formation alteration and yield high-quality paleointensity estimates. In contrast to scenarios of a progressively decaying field leading up to late nucleation of Earth's inner core, these data record a strong field 1.1 Ga. A strong field that persisted over a 14-million-year interval indicates the existence of appreciable power sources for Earth's dynamo during the late Mesoproterozoic Era.}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy due to the exclusion of light elements \cite{Buffett2000a}. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago \cite{Selkin2007a, Biggin2011a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates to maintain the geomagnetic field, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimated thermal conductivity values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates have implied higher thermal conductivity values and an age for the inner core that is less than 1.5 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b, Frost2022a}, with some suggesting even younger ages ($<$700 Ma; \citealp{Labrosse2015a, Ohta2016a, Pozzo2022a}). The possibility of late inner core nucleation has motivated proposals of novel power sources to sustain the geomagnetic field through early Earth history including precipitation of light-element minerals such as MgO \cite{Badro2016a, ORourke2016a, ORourke2016b} and SiO$_2$ \cite{Mittal2020a} at the core-mantle boundary. Given that estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of Earth's core and mantle.

@@

Determinations of the absolute value of ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain the record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluid flow. However, intraplate magmatism associated with large igneous provinces is typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift (Fig. \ref{fig:Geologic_map}) is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma during which there were pulsed intervals of more rapid magmatic activity \cite{Swanson-Hysell2021a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior far from the continental margin and subsequent orogenesis. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and paleomagnetic data from rift rocks form a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it provides an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have posed challenges for the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural remanent magnetization (NRM) is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior due to the presence of nonuniformly magnetized grains (with either multidomain \cite{Dunlop2001a} or vortex states \cite{Tauxe2020a}) is sagging or double-slopes as visualized in Arai plots that show thermal remanent magnetization (TRM) acquired versus NRM lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion of the experiment and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Such non-ideal results were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift. While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions \cite{Swanson-Hysell2021a}. The ca. 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7$\pm$0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can protect magnetic inclusions from alteration. In addition, the alteration of the plagioclase crystals does not readily result in the formation of secondary iron oxides in contrast with Fe-silicates minerals such as olivine and pyroxene. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They were heated to tholeiitic magma temperatures and partially recrystallized \cite{Zhang2021b}. Fe-Ti oxides within the plagioclase exsolved above magnetite Curie temperature ($\sim$580\textdegree C; \citealp{Bian2021a}) and subsequently cooled and acquired thermal remanent magnetizations in conjunction with the host diabase at a paleolatitude of 22 $\pm$ 2\textdegree$\;$(calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a}). Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. The anorthosite xenoliths show low anisotropy of thermal remanent magnetization (TRM) acquisition and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows that the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals without strong preferred orientations. Thermal modeling results and paleomagnetic directional data show that the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths to have dominantly single-component magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with remanence being held by low-titanium titanomagnetite \cite{Zhang2021b}.

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). Plagioclase crystals within coarse-grained intrusions often contain abundant elongate Fe-Ti oxide inclusions that are visible with optical microscopy \cite{Feinberg2006a, Wenk2011a, Ageeva2016a}. The Beaver River anorthosite xenoliths lack such large (10s of $\mu$m) oxide inclusions such that the oxides are not visible within the plagioclase crystals using optical microscopy (Fig. \ref{fig:Petro_QDM}C). Magnetic imaging using a quantum diamond microscope (QDM), however, shows that there are magnetic remanence carriers within the plagioclase crystals (Fig. \ref{fig:Petro_QDM}E, G). The sources of these magnetic remanence carriers are likely to be Fe-Ti oxides that exsolved within plagioclase crystals above the Curie temperature of magnetite \cite{Zhang2021b, Bian2021a}. To highlight these distinctive aspects of the Beaver River anorthosite xenoliths, we present petrographic and magnetic imaging data from the Beaver River anorthosite as well as an anorthosite sample from the Duluth Complex Anorthositic Series---an older intrusive complex within the Midcontinent Rift that was not targeted for paleointensity experiments in this study (Fig. \ref{fig:Petro_QDM}). In contrast to the Beaver River anorthosite, the plagioclase crystals of this Duluth Complex anorthosite sample has a pronounced igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles within the Duluth Complex plagioclase grains that are typically aligned with the [001] axes of the crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these magnetite needles have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}F, H). As a result of this shape anisotropy, remanent magnetizations are acquired at angles highly oblique to applied fields (Figure \ref{fig:Petro_QDM}F). In an experiment where a second orthogonal field was applied, the first isothermal remanent magnetizations (IRM) of a set of magnetite needles within a plagioclase grain either fail to rotate or flip by 180\textdegree. When they flip, they remain in a direction that is oblique to the applied field direction dictated by shape anisotropy (Fig. \ref{fig:Petro_QDM}F, H). These experiments enable novel visualization of the grain-scale magnetic anisotropy of elongated exsolved (titano)magnetite inclusions within plagioclase grains that leads to magnetic anisotropy observed at the bulk sample scale \cite{Selkin2000a, Feinberg2006a}. In contrast, the remanent magnetizations of the ferromagnetic grains imaged in the Beaver River anorthosite xenoliths, targeted for paleointensity in this study, align with the applied field directions indicating minimal remanence anisotropy  (Fig. \ref{fig:Petro_QDM}E, G). The relative lack of petrofabrics and minimal grain-scale magnetic anisotropy make the Beaver River anorthosite xenoliths a particularly compelling target for paleointensity experiments.

@@

Following IZZI-style paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). Seven anorthosite sites and no diabase sites have specimen results that pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1) where each site represents an individual anorthosite xenolith. The paleointensity quality index (Q$_{PI}$; \citealp{Biggin2014a}) for the anorthosite xenolith sites are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from the anorthosite xenoliths have a mean of 38.28$\pm$11.92 $\mu$T. The site-mean virtual dipole moment is $\sim$82 ZAm$^2$ ($10^{21} Am^2$) ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthref.org/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we applied an independent statistical method from ref. \citealp{Cych2021a} to all experimental data regardless of their NRM/TRM plot statistics to make a bias-corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on the assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indicate that the anorthosite xenoliths record a high geomagnetic field ca. 1092 Ma.

\subsection*{Rock magnetism}
\subsubsection*{Coercivity}
Additional rock magnetic data indicate that those anorthosite specimens which pass the paleointensity selection criteria contain dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection along with all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show that both the diabase and anorthosite contain (titano)magnetite as evidenced by the presence of the Verwey transition (Fig. S2;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti contents \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are supressed toward lower temperatures (Fig. S2), indicating that their magnetic grains either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}. Another difference is that samples which pass paleointensity selection have distinctly higher average median destructive field values than other anorthosite and diabase specimens (Fig. \ref{fig:coercivity}). Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites having successful paleointensity results can contain magnetic grain populations with peak coercivities ranging from $\sim$44-144 mT with a median of 53 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase specimens tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that magnetic grain populations with more multidomain-like behavior are responsible for the non-ideal paleointensity behaviors during experiments of such specimens \cite{Xu2004a}.

@@

\subsection*{Considering secular variation and cooling rate}
To best characterize the geomagnetic axial dipole field intensity during a certain time period a paleointensity dataset should cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged. Thermal modeling results from ref. \citealp{Zhang2021b} indicate that the Beaver River anorthosite xenoliths were heated to tholeiitic magma temperature ($\sim$1100\textdegree C)---lower than the melting temperature of the anorthosite plagioclase (given its composition of An$_{70}$; \citealp{Morrison1983a, Doyle2016a}) and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a thousand years, partially averaging secular variation within single sites. Another consequence of the anorthosite xenoliths having slowly cooled in the interior of thick diabase intrusions is that slow cooling rates can bias paleointensity estimates toward higher values \cite{Halgedahl1980a}. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. To fully account for the range of various lower temperature bounds used for optimize the paleointensity estimate and the effect of lowering unblocking temperature due to slow cooling, we estimate the anorthosite cooled over a time period on the order of $\sim$10 kyr (from $\sim$580\textdegree C to as low as $\sim$400\textdegree C) based the thermal history model of ref. \citealp{Zhang2021b} and cooling rate correction model of \cite{Pullaiah1975a}. This cooling time corresponds to a cooling rate of $\sim2.5\times10^{-10}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Pullaiah1975a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. This factor may result in an over-correction, as some multidomain grains and pseudo-single domain grains that do not tend to severely bias paleointensity estimates toward higher \cite{Biggin2013a}, may exist in the anorthosite xenoliths as is evidenced through the slightly zigzagging Arai plots (Figure \ref{fig:IZZI_examples}). The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.
